SUPPLEMENTARY:\\A new approach to bias correction in high-throughput sequencing data

### 2011-11-23 seqbias-paper.tex
Rewordings.

%\title{Correcting for bias in high-throughput sequencing data}
\title{A new approach to bias correction in high-throughput sequencing data}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\hspace{-0.5em}\footnote{to whom correspondence should be addressed}\hspace{0.5em},
Walter L. Ruzzo\,$^{1,2,3}$,
Xinxia Peng$^{4}$
and
Michael G. Katze$^{4}$
}

@@

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated by protocol-specific sequence bias. The exact sources of the
bias are unknown, but may be influenced by PCR amplification, or differing
primer affinities and mixtures, for example.  The result is decreased accuracy
in many applications, such as de novo gene annotation, isoform quantification in
RNA-Seq, and peak calling in ChIP-Seq.

@@

\begin{table}
\processtable{Numbers of reads used to train each method.\label{tab:counts}}
{
\begin{tabular}{lcccc}\toprule
Experiment & Total & 7mer & GLM/MART & BN \\\midrule
Wetterbom & $3.9 \times 10^7$ & $2.3 \times 10^7$ & $5.9 \times
10^6$ & $1 \times 10^5$ \\
Katze & $4.8 \times 10^7$ & $2.0 \times 10^7$ & $3.1 \times 10^6$ & $1 \times
10^5$ \\
Bullard & $1.2 \times 10^7$ & $3.1 \times 10^6$ & $1.8 \times 10^5$ & $1 \times
10^5$\\
Mortazavi & $3.9 \times 10^6$ & $1.9 \times 10^6$ & $3.8 \times 10^5$ & $1
\times 10^5$ \\
Trapnell & $3.4 \times 10^7$ & $1.6 \times 10^7$ & $5.9 \times 10^6$ & $1 \times
10^5$
\end{tabular}
}{Note that in the Trapnell data set (the only paired-end set considered) we
count only the first mate of each read.}
\end{table}

@@

The differences in the methods being tested necessitated training procedures
unique to each. The total number of reads used to train each method is listed in
Table \ref{tab:counts}, and below we describe the procedure used for
each.

@@

To avoid the risk of a small number of reads with many duplicates dominating the
measure, we do not count duplicate reads when computing the KL divergence. Since we
explicitly chose exons with high coverage, not counting duplicates may not
fully capture the effects of the bias correction methods. Thus, for each exon,
we compute the number of duplicates of each read and count only reads in the
upper half. Under the assumption of uniform sampling, the set of reads falling
in the upper half should not depend on sequence, and we should expect the KL
divergence to be low.

We compute the divergence by reweighting the read counts using the predicted
bias coefficient, choosing the upper half in each exon, discarding duplicate
reads, and then tallying frequencies of overlapping $k$-mers. The $k$-mer
distribution obtained is then compared to a background distribution obtained by
redistributing reads uniformly at random within their exons.

We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
remaining cases are plotted in Supplementary Section 3.

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform (see Supplementary Section 5). This evaluation
does not rely on an assumption that qRT-PCR necissarily more accurate than
RNA-Seq based quantification, only that qRT-PCR is not biased in the same way as
the RNA-Seq data.

@@

Training our model on more reads leads to more accurate estimation of bias, but
an increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring less than 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 5.

@@

Additionally, experimental and theoritical analysis suggest that the procedure
is very resistant to inclusion of extranious parameters. In Supplementary
Section 10 we prove an upper bound on the probability of our model predicting
any bias, if the experiment is in fact unbiased, showing that there very little
risk in the applying the method to an unbiased data set. In particular, if more
than 10,000 reads are used in the training procedure, the probability that any
adjustment at all will is made is less than 0.0004.

@@

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. A possible explaination is, given the large number of
parameters needed to estimate heptamer frequencies ($4^{7} = 16,383$), these
parameters are estimated with less accuracy than in models requiring fewer
parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
(see Table \ref{tab:counts}), a number which our theoretical results, detailed
in Supplementary Section 9, suggest should lead to accurate estimates. A perhaps
more significant factor is that this method does not capture bias outside of the
initial heptamer, though many data sets clearly are affected by bias in other
positions.  Thus to improve the performance, it seems necessary to incease the
size of the $k$-mers being considered. Yet, exponentially more reads would be
required for an accurate estimate since the accuracy of the model, as quantified
by its KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is
the number of parameters that must be estimated (See Supplementary Section 9).

@@

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments, may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated; however, our method is still able to effectively correct for the bias
that was observed (see Supplementary Section 6).  Protocol differences, as we
have seen, can result in significant differences in observed nucleotide
frequencies, so we can not safely assert that bias in ChIP-Seq data is always
low.  Given the weak assumptions made by our model, our estimation of bias could
easily be incorporated into ChIP-Seq peak-calling algorithms, and potentially
improve accuracy.

To determine the extent to which PCR amplification is responsible for the
observed bias, we evaluated data from the FRT-Seq method proposed by
\citet{Mamanova2010}.  FRT-Seq avoids the PCR amplification step during library
preparation with reverse transcription occuring on the flowcell surface. We
observed that this data is not free from sequence bias, yet unlike other data
generated on the Illumina platform, it appears to be effected only by relatively
few positions adjacent to the read start (see Supplementary Section 7).

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see Supplementary
Section 8).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
on the Illumina platform display similar patterns of nonuniformity, yet differ
in magnitude, suggesting that batch effects in RNA-Seq remain a legitimate
concern that should not be dismissed without evaluation.

>>>>

%\title{Correcting for bias in high-throughput sequencing data}
\title{A new approach to bias correction in RNA-Seq}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\hspace{-0.5em}\footnote{to whom correspondence should be addressed}\hspace{0.5em},
Walter L. Ruzzo\,$^{1,2,3}$,
Xinxia Peng$^{4}$
and
Michael G. Katze$^{4}$
}

@@

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq experiments is often conflated
by protocol-specific sequence bias. The exact sources of the bias are unknown,
but may be influenced by PCR amplification, or differing primer affinities and
mixtures, for example. The result is decreased accuracy in many applications,
such as de novo gene annotation and gene or isoform quantification.

@@

@@

The differences in the methods being tested necessitated training procedures
unique to each. The total number of reads used to train each method is listed in
Supplementary Section 3, and below we describe the procedure used for
each.

@@

When computing the KL divergence, there is a risk of the measure being dominated
by a small number of reads with many duplicates. Yet, given the high coverage of
the exons being tested, if duplicate reads are excluded, it may not capture the
full effect of bias correction. To account for these opposing concerns we adopt
the following method: all reads contained within the exon being tested are
ranked by the number of duplicates. We then exclude reads that are ranked in the
lower half, and count each read ranked in the upper half only once, ignoring
duplicates.

Under the assumption of uniform sampling, the set of reads ranked in the upper
half should not depend on sequence, and we should expect the KL divergence to be
low.  We compute the divergence by reweighting the read counts using the
predicted bias coefficient before ranking the reads, choosing those reads
ranked in the upper half of each exon, ignoring duplicate reads, and then
tallying frequencies of overlapping $k$-mers. The $k$-mer distribution obtained
is then compared to a background distribution obtained by redistributing reads
uniformly at random within their exons.

We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
remaining cases are plotted in Supplementary Section 4.

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform (see Supplementary Section 6). This evaluation
does not rely on an assumption that qRT-PCR necissarily more accurate than
RNA-Seq based quantification, only that qRT-PCR is not biased in the same way as
the RNA-Seq data.

@@

Training our model on more reads leads to more accurate estimation of bias, but
an increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring less than 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 6.

@@

Additionally, experimental and theoritical analysis suggest that the procedure
is very resistant to inclusion of extranious parameters. In Supplementary
Section 11 we prove an upper bound on the probability of our model predicting
any bias, if the experiment is in fact unbiased, showing that there very little
risk in the applying the method to an unbiased data set. In particular, if more
than 10,000 reads are used in the training procedure, the probability that any
adjustment at all will is made is less than 0.0004.

@@

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. A possible explaination is, given the large number of
parameters needed to estimate heptamer frequencies ($4^{7} = 16,383$), these
parameters are estimated with less accuracy than in models requiring fewer
parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
(see Supplementary Section 3), a number which our theoretical results, detailed
in Supplementary Section 10, suggest should lead to accurate estimates. A perhaps
more significant factor is that this method does not capture bias outside of the
initial heptamer, though many data sets clearly are affected by bias in other
positions.  Thus to improve the performance, it seems necessary to incease the
size of the $k$-mers being considered. Yet, exponentially more reads would be
required for an accurate estimate since the accuracy of the model, as quantified
by its KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is
the number of parameters that must be estimated (See Supplementary Section 10).

@@

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments, may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated; however, our method is still able to effectively correct for the bias
that was observed (see Supplementary Section 7).  Protocol differences, as we
have seen, can result in significant differences in observed nucleotide
frequencies, so we can not safely assert that bias in ChIP-Seq data is always
low.  Given the weak assumptions made by our model, our estimation of bias could
easily be incorporated into ChIP-Seq peak-calling algorithms, and potentially
improve accuracy.

To determine the extent to which PCR amplification is responsible for the
observed bias, we evaluated data from the FRT-Seq method proposed by
\citet{Mamanova2010}.  FRT-Seq avoids the PCR amplification step during library
preparation with reverse transcription occuring on the flowcell surface. We
observed that this data is not free from sequence bias, yet unlike other data
generated on the Illumina platform, it appears to be effected only by relatively
few positions adjacent to the read start (see Supplementary Section 8).

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see Supplementary
Section 9).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
on the Illumina platform display similar patterns of nonuniformity, yet differ
in magnitude, suggesting that batch effects in RNA-Seq remain a legitimate
concern that should not be dismissed without evaluation.

### 2011-11-23 seqbias-supplement.tex
Rewordings.

>>>>

\section{Total numbers of reads used to train each method}

\begin{table}[H]
\begin{center}
\begin{tabular}{lcccc}
Experiment & Total & 7mer & GLM/MART & BN \\\hline
Wetterbom & $3.9 \times 10^7$ & $2.3 \times 10^7$ & $5.9 \times
10^6$ & $1 \times 10^5$ \\
Katze & $4.8 \times 10^7$ & $2.0 \times 10^7$ & $3.1 \times 10^6$ & $1 \times
10^5$ \\
Bullard & $1.2 \times 10^7$ & $3.1 \times 10^6$ & $1.8 \times 10^5$ & $1 \times
10^5$\\
Mortazavi & $3.9 \times 10^6$ & $1.9 \times 10^6$ & $3.8 \times 10^5$ & $1
\times 10^5$ \\
Trapnell & $3.4 \times 10^7$ & $1.6 \times 10^7$ & $5.9 \times 10^6$ & $1 \times
10^5$
\end{tabular}
\caption{Note that in the Trapnell data set (the only paired-end set considered) we
count only the first mate of each read.}
\label{tab:counts}
\end{center}
\end{table}
